---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# diverstidy: A tidy package for detecting gender, race, and other forms of diversity in unstandardized text data

**Authors:** [Brandon Kramer](https://www.brandonleekramer.com/)<br/>
**License:** [MIT](https://opensource.org/licenses/MIT)<br/>

### Installation

You can install this package using the `devtools` package:

```{r, eval=FALSE, warning=FALSE, message=FALSE}
install.packages("devtools")
devtools::install_github("brandonleekramer/diverstidy") 
```

The `diverstidy` package provides several functions that help detect patterns in unstandardized text data for analyses of geographies, populations, other forms of diversity. Currently, there are 17 different functions that detect terms across the following subdomains of diversity-related research: ancestry, culture, disability, discrimination, diversity, equity, inclusion, linguistic, migration, population, race/ethnicity, religious, sex/gender, sexuality, social class, and US OMB population terms. Although somewhat simple, the intuition behind these functions is to detect the quantity of diversity-related terms show up in a given text entry. To do this, each function depends on a curated dictionary of terms that fall under these 17 domains of topics, which can be called using the `data(diversity_dictionary)` function. There are a number of case studies, but the primary uses of these functions are to examine historical trends in term usage and/or to detect potential biases in text. 

### Detecting racial/ethnic terms with the `detect_racialethnic_terms()` function

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(diverstidy)
data(pubmed_data)
```

```{r, warning=FALSE, message=FALSE, fig.align="center"}
pubmed_data %>%
  detect_racialethnic_terms(fk_pmid, abstract) %>%
  detect_sexgender_terms(fk_pmid, abstract) %>% 
  detect_socialclass_terms(fk_pmid, abstract) %>% 
  group_by(year) %>% 
  summarize(racial_ethnic = sum(racial_ethnic),
            sex_gender = sum(sex_gender),
            social_class = sum(social_class)) %>% 
  pivot_longer(!year, names_to = "category", values_to = "count") %>% 
  ggplot(aes(x=year, y=count, group=category)) +
  geom_line(aes(color=category), size = 1) +
  theme_bw() + 
  ggtitle("Change in Diversity-Related Terms Over Time")
```

### Using the `diversity_dictionary` to examine relationships within text 

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(diverstidy)
library(tidytext)
library(igraph)
library(ggraph)
library(tidygraph)
data(pubmed_data)
data(diversity_dictionary)
```

```{r, warning=FALSE, message=FALSE, fig.align="center"}
# create an edgelist of all terms mentioned more than 100 times together 
pubmed_graph <- pubmed_data %>%
  unnest_tokens(bigram, abstract, token = "ngrams", n = 2) %>%
  separate(bigram, c("word1", "word2"), sep = " ") %>% 
  count(word1, word2, sort = TRUE) %>%
  filter(n > 100) %>%
  graph_from_data_frame() 

# pull out the nodelist from that graph 
nodelist <- data.frame(id = c(1:(igraph::vcount(pubmed_graph))), 
                       name = igraph::V(pubmed_graph)$name)
# and join category data to all our diversity terms 
dictionary_terms <- diversity_dictionary %>% 
  unnest_legacy(name = strsplit(catch_terms, "\\|")) %>% 
  select(name, category)
nodelist <- nodelist %>% 
  left_join(dictionary_terms, by = "name") %>% 
  mutate(category = replace_na(category, "nothing"),
         category = str_replace(category, "race/ethnicity\\|us omb terms", "us omb terms"))
V(pubmed_graph)$category <- nodelist$category

# create a custom color palette and vector to only visualize certain words 
custom_colors <- colorRampPalette(c("#D3D3D3", RColorBrewer::brewer.pal(9, 'Spectral')))
graph_tbl <- pubmed_graph %>% 
  as_tbl_graph() %>% 
  activate(nodes) %>% 
  mutate(degree  = centrality_degree()) %>% 
  mutate(new_name = ifelse(str_detect(
    name, str_c("\\b(?i)(",paste0(dictionary_terms$name, collapse = "|"),")\\b")), name, no = ""))

# graph a text network of all our dictionary terms 
layout <- create_layout(graph_tbl, layout = 'igraph', algorithm = 'nicely')
ggraph(layout) +
  geom_edge_fan(aes(alpha = ..index..), show.legend = F) + 
  geom_node_point(aes(size = degree, color = as.factor(category)), show.legend = F) +
  geom_node_text(aes(label = new_name), vjust = 1, hjust = 1) +
  scale_color_manual(limits = as.factor(layout$category), 
                     values = custom_colors(nrow(layout))) +  
  theme_void()
```





